\documentclass[table,svgnames]{beamer}

\input{packages}
\input{config}


\title{\Large{Integrating Heterogeneous Data by Extending Semantic Web Standards}}

\author{Nuno Lopes}

\date{26 November, 2012}


\begin{document}

\begin{frame}[plain]
  \titlepage
\end{frame}


\section{Overview}
\input{motivating-usecase}

{
  \setbeamertemplate{footline}{}
  \addtocounter{framenumber}{-1}
  \begin{frame}{Outline and Contributions} 
    \begin{tikzpicture}
      \node[text width=.5\linewidth] (two) {
        \begin{itemize}
        \item XSPARQL
          \begin{itemize}
          \item Syntax \& Semantics
          \item Implementation
          \item Evaluation 
          \end{itemize}
        \end{itemize}
      };
      \node[box=green,right=of two,visible on=<2>,text width=3cm] {\cite{BischofDeckerKrennwallner:2012aa}, \cite{LopesBischofDecker2011aa}};
      \node[text width=.5\linewidth,below = 1em of two] (three){
        \begin{itemize}
        \item Annotated RDF(S)
          \begin{itemize}
          \item Annotated RDF(S)
          \item AnQL: Annotated SPARQL
          \item Architecture
          \end{itemize}
        \end{itemize}
      };
      \node[box=green,right=of three,visible on=<3>,text width=3cm] {\cite{StracciaLopesLukacsy:2010aa}, \cite{LopesPolleresStraccia:2010aa}, \cite{ZimmermannLopesPolleres:2012aa}};
    \end{tikzpicture}
  \end{frame}
}

\section{XSPARQL}
\input{1-xsparql/xsparql}

\section{Annotated RDF(S)}
\input{2-anql/anql}

\input{3-combination/combining-xsparql-anql}

\input{conclusions}

\newcounter{finalframe}
\setcounter{finalframe}{\value{framenumber}}

\appendix

\section*{Bibliography}
{
  \setbeamertemplate{footline}{}
  \addtocounter{framenumber}{-1}
  \begin{frame}[allowframebreaks]{Bibliography}
    \def\newblock{}
\begin{thebibliography}{BDK{\etalchar{+}}12}

% \bibitem[JoDS2012]{BischofDeckerKrennwallner:2012aa}
\bibitem[JoDS2012]{BischofDeckerKrennwallner:2012aa}
Stefan Bischof, Stefan Decker, Thomas Krennwallner, Nuno Lopes, and Axel
  Polleres.
\newblock {Mapping between RDF and XML with XSPARQL}.
\newblock {\em Journal on Data Semantics}, 1:147--185, 2012.

% \bibitem[GGL{\etalchar{+}}08]{GroppeGroppeLinnemann:2008aa}
\bibitem[SAC2008]{GroppeGroppeLinnemann:2008aa}
Sven Groppe, Jinghua Groppe, Volker Linnemann, Dirk Kukulenz, Nils Hoeller, and
  Christoph Reinke.
\newblock {Embedding SPARQL into XQuery/XSLT}.
\newblock In Roger~L. Wainwright and Hisham Haddad, editors, {\em {Proceedings
  of the 2008 ACM Symposium on Applied Computing (SAC), Fortaleza, Ceara,
  Brazil, March 16-20, 2008}}, pages 2271--2278. ACM, 2008.

% \bibitem[LBDP11]{LopesBischofDecker2011aa}
\bibitem[EPIA2011]{LopesBischofDecker2011aa}
Nuno Lopes, Stefan Bischof, Stefan Decker, and Axel Polleres.
\newblock {On the Semantics of Heterogeneous Querying of Relational, XML and
  RDF Data with XSPARQL}.
\newblock In Paulo Moura and Vitor~Beires Nogueira, editors, {\em {Proceedings
  of the 15th Portuguese Conference on Artificial Intelligence (EPIA2011) --
  Computational Logic with Applications Track}}, Lisbon, Portugal, October
  2011.

% \bibitem[LKZ{\etalchar{+}}12]{LopesKirraneZimmermann:2012aa}
\bibitem[ICLP2012]{LopesKirraneZimmermann:2012aa}
Nuno Lopes, Sabrina Kirrane, Antoine Zimmermann, Axel Polleres, and Alessandra
  Mileo.
\newblock {A Logic Programming approach for Access Control over RDF}.
\newblock In Agostino Dovier and V{\'\i}tor~Santos Costa, editors, {\em
  {Technical Communications of the 28th International Conference on Logic
  Programming (ICLP'12)}}, volume~17 of {\em Leibniz International Proceedings
  in Informatics (LIPIcs)}, pages 381--392, Dagstuhl, Germany, 2012. Schloss
  Dagstuhl--Leibniz-Zentrum fuer Informatik.

\bibitem[ISWC2010]{LopesPolleresStraccia:2010aa}
Nuno Lopes, Axel Polleres, Umberto Straccia, and Antoine Zimmermann.
\newblock {AnQL: SPARQLing Up Annotated RDFS}.
\newblock In {\em {International Semantic Web Conference (1)}}, pages 518--533,
  2010.

\bibitem[AAAI2010]{StracciaLopesLukacsy:2010aa}
Umberto Straccia, Nuno Lopes, Gergely Luk{\'a}csy, and Axel Polleres.
\newblock {A General Framework for Representing and Reasoning with Annotated
  Semantic Web Data}.
\newblock In Maria Fox and David Poole, editors, {\em {Proceedings of the
  Twenty-Fourth AAAI Conference on Artificial Intelligence, AAAI 2010, Atlanta,
  Georgia, USA, July 11-15, 2010}}. AAAI Press, July 2010.

\bibitem[JWS2012]{ZimmermannLopesPolleres:2012aa}
Antoine Zimmermann, Nuno Lopes, Axel Polleres, and Umberto Straccia.
\newblock {A General Framework for Representing, Reasoning and Querying with
  Annotated Semantic Web Data}.
\newblock {\em Web Semantics: Science, Services and Agents on the World Wide
  Web}, 11(0):72--95, 2012.

\end{thebibliography}

  \end{frame}
}


\setcounter{framenumber}{\value{finalframe}}

\end{document}

%%% Local Variables:
%%% mode: latex
%%% mode: flyspell
%%% mode: reftex
%%% End:
