\section{RDF(S) with Annotations}
\label{sec:annotated-rdfs}


For extending our running example we use the temporal domain, which allows us to annotate the RDF data with temporal
information. For instance, we can annotate the band members' triples to reflect their active years with the band. 
%
A possible temporal query that can be easily performed over data represented in this format is ``What were the members
of a band at a certain time?''
%
The use of other domains would allow to represent different views on the data, for example in the fuzzy domain, we can
represent information regarding part-time members of bands.



\begin{data}[t]
  \centering
  \lstinputlisting[basicstyle=\ttfamily\footnotesize]{0-data+queries/usecasedata-temporalRDF.ttl}
  %
  \caption{Temporal Annotated RDFS}
  \label{fig:dataset-example}
\end{data}



\cref{fig:dataset-example} represents an extension of our \usecase data from Data~\ref{fig:bands-turtle} annotated
with information from the temporal domain, which intuitively means that the annotated triple is valid in dates contained
in the annotation interval (the exact meaning of the annotations will be explained later).
%
In \cref{fig:dataset-example}, we are representing the annotated triples using N-Quads~\cite{CyganiakHarthHogan:2009aa},
a format that allows to attach a forth element to an \ac{RDF} triple.\footnote{A similar approach is followed for
  extending SPARQL's syntax.}
%
However, in examples and definitions of the rest of this chapter we will use a representation of the
form~$\fuzzyg{\triple{s,p,o}}{\lambda}$, which is considered equivalent to its N-Quad counterpart. 





\newif\ifnormalisedardf\normalisedardftrue
% \normalisedardffalse

\subsection{Syntax}

Our approach is to extend triples with annotations, where an annotation is taken from a specific domain.
%
This extension follows a similar approach to the annotated logic programming framework~\cite{KiferSubrahmanian:1992aa}.

\ifnormalisedardf%
\begin{definition}[Annotated RDF triple and graph]
  An \emph{annotated triple} is an expression~$\fuzzyg{\tau}{\lambda}$, where~$\tau$ is an \ac{RDF} triple and~$\lambda$
  is an \emph{annotation value} (defined below). An \emph{annotated graph} is a finite set of annotated triples.
  % 
  Furthermore we call an annotated graph~$G$ a \emph{normalised annotated graph}
  iff~$\not\exists\fuzzyg{\tau}{\lambda_{1}},\fuzzyg{\tau}{\lambda_{2}} \in G$ s.t.~$\lambda_{1}\neq\lambda_{2}$.
\end{definition}
% 
\else%
An \emph{annotated triple} is an expression~$\fuzzyg{\tau}{\lambda}$, where~$\tau$ is an \ac{RDF} triple and~$\lambda$
is an \emph{annotation value} (defined below). An \emph{annotated graph} is a finite set of annotated triples.
% 
\fi
%
The intended semantics of annotated triples depends of course on the meaning we associate to the annotation values. For
instance, in a temporal setting~\cite{GutierrezHurtadoVaisman:2007aa},
\[
\fuzzyg{\triple{\qname{dbpedia}{Nightwish}, \qname{foaf}{member}, \qname{dbpedia}{Marco\mathunderscore{}Hietala}}}{{[2001,2012]}}
\]
\nd has the intended meaning ``Marco Hietala was a member of Nightwish during the period from~$2001$ to~$2012$'',
% 
while in the fuzzy setting~\cite{Straccia:2009aa}, we can represent part-time members of a band:
\[
\fuzzyg{\triple{\qname{dbpedia}{Nightwish}, \qname{foaf}{member}, \qname{dbpedia}{Troy\mathunderscore{}Donockley}}}{0.7}
\]
\nd with the intended meaning ``Troy Donockley is a member of Nightwish to a degree not less than 0.7''.\footnote{The
  membership degree was chosen as an example, Troy has collaborated with Nightwish on different albums and live
  concerts.}


\subsection{Annotation Domain Specification}
\label{sec:rdfs-annot-doma}



To start with, let us consider a non-empty set~$L$, where the elements in~$L$ are our annotation values. For example, in
a fuzzy setting,~$L = [0,1]$, while in a typical temporal setting,~$L$ may be time points or time intervals.  In our
annotation framework, we extend the notion of interpretation (presented in \cref{def:interpretation}) to map
statements to elements of the annotation domain.  
%
But first let us define an annotation domain:




\begin{definition}[Annotation Domain]
  We say that an \emph{annotation domain} for \ac{RDFS} is an idempotent, commutative semi-ring
  \[
  D = \tuple{L,  \oplus, \otimes, \bot, \top} \ ,
  \]
  \nd where,~$\top,\bot$ are specific annotation values and~$\oplus$
  is~$\top$-annihilating~\citep{BunemanKostylev:2010aa}. That is, for~$\lambda, \lambda_{i} \in L$:

  \begin{enumerate}[noitemsep]
  \item $\oplus$ is idempotent, commutative, associative;
  \item $\otimes$ is commutative and associative;
  \item $\bot \oplus \lambda = \lambda$,~$\top \otimes \lambda = \lambda$,~$\bot \otimes \lambda = \bot$, and~$\top \oplus \lambda = \top$;
  \item $\otimes$ is distributive over~$\oplus$, \ie~$\lambda_{1} \otimes (\lambda_{2} \oplus \lambda_{3}) =
    (\lambda_{1} \otimes \lambda_{2}) \oplus (\lambda_{1} \otimes \lambda_{3})$.
  \end{enumerate}
\end{definition}
%
\nd Please note that there is a natural partial order on any idempotent semi-ring: an annotation domain~$D = \tuple{L,
  \oplus, \otimes, \bot, \top}$ induces a partial order~$\preceq$ over~$L$ defined as:
%
\[
\lambda_{1} \preceq \lambda_{2} \mbox{ \ \iff  \ }  \lambda_{1} \oplus \lambda_{2} = \lambda_{2} \ .
\]
%
The~$\top$ and~$\bot$ respective represent the highest and lowest element in the partial order.  This partial
order~$\preceq$ is used to express redundant information: for instance, for temporal intervals, an annotated
triple~$\fuzzyg{\triple{s,p,o}}{[2000,2006]}$ includes~$\fuzzyg{\triple{s,p,o}}{[2003,2004]}$, since~$[2003,2004]
\subseteq [2000,2006]$ (here,~$\subseteq$ plays the role of~$\preceq$).


In previous work~\cite{StracciaLopesLukacsy:2010aa,LopesPolleresStraccia:2010aa}, an annotation domain was assumed to be
a more specific structure, namely a residuated bounded lattice.  In~\citet{BunemanKostylev:2010aa} it was shown that we
may use a slightly weaker structure than residuated lattices for annotation domains.




We use~$\oplus$ to combine information about the same statement.  For instance, in temporal logic,
from~$\fuzzyg{\tau}{[2000,2006]}$ and~$\fuzzyg{\tau}{[2003,2008]}$, we infer~$\fuzzyg{\tau}{[2000,2008]}$,
as~$[2000,2008] = [2000,2006]\cup [2003,2008]$ (where~$\cup$ plays the role of~$\oplus$).
%
In the fuzzy context, from~$\fuzzyg{\tau}{0.7}$ and~$\fuzzyg{\tau}{0.6}$, we infer~$\fuzzyg{\tau}{0.7}$, since~$0.7 =
\max(0.7, 0.6)$ (here,~$\max$ plays the role of~$\oplus$).

We use~$\otimes$ to model the ``conjunction'' of information, where~$\otimes$ is a generalisation of boolean conjunction
to the many-valued case. In fact,~$\otimes$ satisfies also that:
%
\begin{enumerate}[noitemsep]
\item $\otimes$ is bounded: \ie~$\lambda_{1} \otimes \lambda_{2} \preceq \lambda_{1}$.
\item $\otimes$ is~$\preceq$-monotone, \ie~for~$\lambda_{1} \preceq \lambda_{2}$,~$\lambda \otimes \lambda_{1} \preceq
  \lambda \otimes \lambda_{2}$
\end{enumerate}
%
For instance, on interval-valued temporal logic, from~$\fuzzyg{\triple{a, \subclass, b}}{[2000,2006]}$ and
$\fuzzyg{\triple{b, \subclass, c}}{[2003,2008]}$, we will infer~$\fuzzyg{\triple{a, \subclass, c}}{[2003,2006]}$, as
$[2003,2006] = [2000,2006] \cap [2003,2008]$ (where~$\cap$ plays the role of~$\otimes$).
%
In the fuzzy context, one may chose any triangular norm (t-norm)~\cite{KlementMesiarPap:2000aa}, \eg~product, and, thus,
from~$\fuzzyg{\triple{a, \subclass, b}}{0.7}$ and~$\fuzzyg{\triple{b, \subclass, c}}{0.6}$, we will
infer~$\fuzzyg{\triple{a, \subclass, c}}{0.42}$, as $0.42 = 0.7 \cdot 0.6$ (here,~$\cdot$ plays the role of~$\otimes$).

The distributivity condition guarantees that we obtain the same annotation~$\lambda_{1} \otimes (\lambda_{2} \oplus
\lambda_{3}) = (\lambda_{1} \otimes \lambda_{2}) \oplus (\lambda_{1} \otimes \lambda_{3})$ of the triple~$\triple{a,
  \subclass, c}$ that can be inferred from triples~$\fuzzyg{\triple{a, \subclass, b}}{\lambda_{1}}$, $\fuzzyg{\triple{b,
    \subclass, c}}{\lambda_{2}}$ and~$\fuzzyg{\triple{b, \subclass, c}}{\lambda_{3}}$.
%
Finally, note that, conceptually, in order to build an annotation domain, one has to:
% 
\begin{enumerate}[noitemsep]
\item determine the set of annotation values~$L$ (typically a countable set\footnote{Note that one may use \ac{XML}
    decimals in~$[0,1]$ in place of real numbers for the fuzzy domain.}),
  identifying the top and bottom elements;

\item define suitable operations~$\otimes$ and~$\oplus$ that acts as ``conjunction'' and ``disjunction'' functions, to
  support the intended inference over schema axioms, such as
  \begin{quote}
    ``from~$\fuzzyg{\triple{a, \subclass, b}}{\lambda}$ and~$\fuzzyg{\triple{b, \subclass,
        c}}{\lambda'}$ infer~$\fuzzyg{\triple{a, \subclass, c}}{\lambda \otimes \lambda'}$''
  \end{quote}

  \nd and
  \begin{quote}
    ``from~$\fuzzyg{\tau}{\lambda}$ and~$\fuzzyg{\tau}{\lambda'}$ infer~$\fuzzyg{\tau}{\lambda \oplus \lambda'}$''
  \end{quote}
\end{enumerate}

Another desirable feature is to use annotated and non-annotated triples in parallel, possibly even in the same dataset.
In~\citet{ZimmermannLopesPolleres:2012aa}, we presented several approaches for combining annotated and non-annotated
triples, such as assuming a default annotation for any non-annotated triple or creating a new compound domain.
% 
For simplicity, and since we are considering the issue of compound domains as out of scope for this thesis, we follow
the approach of assuming a default annotation for non-annotated triples. 
%
This default annotation can be specified on a per-domain basis however, if unspecified, we assume the~$\top$ element
from the domain as the default annotation.





\subsection{Semantics}
\label{sec:anql-semantics}
%
For this section, we fix an annotation domain~$D = \tuple{L, \oplus, \otimes, \bot, \top}$.
%
Similar to \cref{sec:rdf-semantics} we rely on the \rhodf fragment of \ac{RDFS} and do not consider datatype
interpretations.
%
\begin{definition}[Annotated Map]
  An \emph{annotated map} is a function~$\theta : \AUBL \to \AUBL$ preserving \acp{URI} and literals,
  \ie~$\fc{\theta}{t} = t$, for all~$t \in \AUL$.
  % 
  Given an annotated graph~$G$, we define~$\fc{\theta}{G} = \set{ \fuzzyg{\triple{\fc{\theta}{s}, \fc{\theta}{p},
        \fc{\theta}{o}}}{\lambda'} \mid \fuzzyg{\triple{s, p, o}}{\lambda} \in G }$, where~$\lambda' \in \AL$
  and~$\lambda' \preceq \lambda$.
  %
  Similarly to the classical case, we speak of an annotated map~$\theta$ from~$G_{1}$ to~$G_{2}$, and write~$\theta :
  G_{1} \to G_{2}$, if~$\theta$ is such that~$\forall \fuzzyg{\tau}{\lambda_{2}} \in G_{2},
  \exists\fuzzyg{\tau}{\lambda_{1}} \in G_{1}$ such that~$\lambda_{2} \preceq \lambda_{1}$.\footnote{As a shorthand
    notation, from herein we will use~$G_{2} \preceq G_{1}$ to denote~$\forall \fuzzyg{\tau}{\lambda_{2}} \in G_{2},
    \exists\fuzzyg{\tau}{\lambda_{1}} \in G_{1}$ such that~$\lambda_{2} \preceq \lambda_{1}$.}
  %
\end{definition}
%
Informally, an interpretation~$\I$ will assign to a triple~$\tau$ an element of the annotation domain~$\lambda \in L$:
%
\begin{definition}[Annotated Interpretation, extends \cref{def:interpretation}]
  An \emph{annotated interpretation}~$\I$ over a vocabulary~$V$ is a tuple
  % 
  \[\I=\tuple{\Delta_{R}, \Delta_{P}, \Delta_{C}, \Delta_{L}, \intP{\cdot}, \intC{\cdot}, \int{\cdot}}\]
  % 
  where~$\Delta_{R}, \Delta_{P}, \Delta_{C}, \Delta_{L}$ are interpretation domains of~$\I$ and~$\intP{\cdot},
  \intC{\cdot}, \int{\cdot}$ are interpretation functions of~$\I$.
  %
  They have to satisfy:
  \begin{enumerate}
  \item $\Delta_{R}$ is a nonempty finite set of resources, called the domain or universe of~$\I$;
  \item $\Delta_{P}$ is a finite set of property names (not necessarily disjoint from~$\Delta_{R}$);
  \item $\Delta_{C} \subseteq \Delta_{R}$ is a distinguished subset of~$\Delta_{R}$ identifying if a resource denotes a
    class of resources;
  \item $\Delta_{L} \subseteq \Delta_{R}$, the set of literal values,~$\Delta_{L}$ contains all plain literals in~$\AL
    \cap V$;
  \item\label{item5} $\intP{\cdot}$ maps each property name~$p \in \Delta_{P}$ into a function~$\intP{p}: \Delta_{R}
    \times \Delta_{R}\to L$, \ie~assigns an annotation value to each pair of resources;
  \item\label{item6} $\intC{\cdot}$ maps each class~$c \in \Delta_{C}$ into a function~$\intC{c} : \Delta_{R} \to L$,
    \ie~assigns an annotation value representing class membership in~$c$ to every resource;
  \item $\int{\cdot}$ maps each~$t \in \AUL \cap V$ into a value~$\int{t} \in \Delta_{R} \cup \Delta_{P}$ and such
    that~$\int{\cdot}$ is the identity for plain literals and assigns an element in~$\Delta_{R}$ to each element
    in~$\AL$.
  \end{enumerate}
\end{definition}


Similar to the classical case we provide a single notion of interpretation that covers Simple, \ac{RDF}, and \ac{RDFS}
(\rhodf) entailment.  Furthermore, we extend the definition of model:
%
\begin{definition}[Model, extends \cref{def:rdf-model}]
  \label{def:annotated-model}%
  \nd An interpretation~$\I$ is a \emph{model} of an annotated ground graph~$G$, denoted~$\I \models G$, \iff~$\I$ is an
  interpretation over the vocabulary~$\rhodf \cup \fc{\universe}{G}$ that satisfies the following conditions, where~$A
  \colon \AB \maps \Delta_{R}$ and~$\I_{A}$ is the extension of~$\I$ with~$A$:
  %
  \begin{description}
  \item[Simple:] \
    \begin{enumerate}
    \item $\fuzzyg{\triple{s, p, o}}{\lambda} \in G$ implies~$\intA{p} \in\Delta_{P}$ and~$\intP{\intA{p}}(\intA{s}, \intA{o})
      \succeq \lambda$;
    \end{enumerate}
  \item[Subproperty:] \
    \begin{enumerate}
    \item $\intP{\int{ \spp}}(p, q) \otimes \intP{\int{ \spp}}(q, r) \preceq \intP{\int{ \spp}}(p, r)$;
    \item $\intP{\int{ p}}(x, y) \otimes \intP{\int{ \spp}}(p, q) \preceq \intP{\int{ q}}(x, y)$;
    \end{enumerate}
  \item[Subclass:] \
    \begin{enumerate}
    \item $\intP{\int{ \subclass}}(c, d) \otimes \intP{\int{ \subclass}}(d, e) \preceq \intP{\int{ \subclass}}(c, e)$;
    \item $\intC{\int{c}}(x) \otimes \intP{\int{ \subclass}}(c, d) \preceq \intC{\int{ d}}(x)$;
    \end{enumerate}
  \item[Typing I:] \
    \begin{enumerate}
    \item $\intC{c}(x) =  \intP{\int{\typeR}}(x,c)$;
    \item $\intP{\int{\domR}}(p, c) \otimes  \intP{p}(x,y) \preceq  \intC{c}(x)$;
    \item $\intP{\int{\range}}(p, c) \otimes  \intP{p}(x,y) \preceq  \intC{c}(y)$;
    \end{enumerate}
  \item[Typing II:] \
    \begin{enumerate}
    \item For each~$\eee \in \rhodf$,~$\int{\eee} \in \Delta_{P}$; 
    \item $\intP{\int{\spp}}(p, q)$ is defined only for~$p,q \in \Delta_{P}$;
    \item $\intC{\int{\subclass}}(c, d)$ is defined only for~$c,d \in \Delta_{C}$;
    \item $\intP{\int{\domR}}(p, c)$ is defined only for~$p \in \Delta_{P}$ and~$c \in \Delta_{C}$;
    \item  $\intP{\int{\range}}(p, c)$ is defined only for~$p \in \Delta_{P}$ and~$c \in \Delta_{C}$;
    \item $\intP{\int{\typeR}}(s, c)$ is defined only for~$c \in \Delta_{C}$.
    \end{enumerate}
  \end{description}
\end{definition}
%
\nd Intuitively, a triple~$\fuzzyg{\triple{s, p, o}}{\lambda}$ is satisfied by $\I$ if~$(s, o)$ belongs to the extension
of~$p$ to a ``wider'' extent than~$\lambda$.
%
Note that the major differences from the classical setting reside on \cref{item5,item6}.


Finally, entailment among annotated ground graphs~$G$ and~$H$ is as usual.
Now,~$G \models H$, where~$G$ and~$H$ may contain blank nodes, \iff for any grounding~$G'$ of~$G$ there is a grounding~$H'$ of~$H$ such that~$G' \models H'$.

Please note that we always have that~$G \models \fuzzyg{\tau}{\bot}$, however, triples of the form~$\fuzzyg{\tau}{\bot}$
are uninteresting and, thus, in the following we do not consider them as part of the language.


\subsection{Examples of Annotation Domains}
\label{sec:domain-examples}


Next we specify some domains in Annotated RDFS, namely the classical domain, fuzzy~\cite{Straccia:2009aa},
temporal~\cite{GutierrezHurtadoVaisman:2007aa}, and provenance.

\subsubsection{The Classical Domain}
\label{sec:crisp}

The classical \ac{RDF} setting corresponds to the case in which the annotation values are~$L = \set{0,1}$.
%
Thus, the classical domain can be specified as~$D_{01} = \tuple{\set{0, 1}, \max, \min, 0, 1}$.  In this case, Annotated
RDFS turns out to be the same as standard \ac{RDFS}.


\subsubsection{The Temporal Domain}
\label{sec:temporal}


For our representation of the temporal domain we aim at using non-discrete time as it is necessary to model temporal
intervals with any precision.
% 
However, for presentation purposes we will show the dates as years only.
%
To start with, \emph{time points} are elements of the value space~$\mathbb{Q} \cup\ \set{-\infty, +\infty}$ and a
\emph{temporal interval} is a non-empty interval~$[\alpha_1, \alpha_2]$, where $\alpha_{i}$ are time points.  An empty
interval is denoted as~$\emptyset$.
%
We define the partial order on intervals as~$I_{1} \leq I_{2} \mbox{ \iff\ } I_{1} \subseteq I_{2}$.  The intuition here
is that if a triple is true at time points in~$I_{2}$ and~$I_{1} \leq I_{2}$ then, in particular, it is true at any time
point in~$I_{1} \neq \emptyset$.

Now, apparently the set of intervals would be a candidate for~$L$, however, in order to represent the upper bound
interval of~$\fuzzyg{\tau}{[2001,2005]}$ and~$\fuzzyg{\tau}{[2008,2009]}$ we rather need the union of intervals,
denoted~$\set{[2001,2005], [2008,2009]}$, meaning that a triple is true both in the former as well as in the latter
interval.
%
Now, we define~$L$ as:
%
\[
L = \set{t \mid t \mbox{ is a finite set of disjoint temporal intervals} } \cup \set{\bot,\top} \enspace ,
\]
%
where~$\bot = \{\emptyset\}, \top = \{[-\infty,+\infty]\}$.
%
Therefore, a \emph{temporal term} is an element~$t \in L$, \ie~a set of pairwise disjoint time intervals. 
% 
We allow the following variations:
\begin{enumerate}[noitemsep,label=(\roman*)]
\item $[\alpha]$ as a shorthand for~$[\alpha, \alpha]$;
\item $\fuzzyg{\tau}{\alpha}$ as a shorthand of $\fuzzyg{\tau}{\{[\alpha]\}}$; and
\item $\fuzzyg{\tau}{[\alpha, \alpha']}$ as a shorthand of~$\fuzzyg{\tau}{\{[\alpha, \alpha']\}}$.
\end{enumerate}
%
Furthermore, on~$L$ we define the following partial order:
\[
t_1 \preceq t_2 \mbox{ \iff }~ \forall I_{1} \in t_1 \exists I_2 \in t_2, \mbox{ such that } I_1 \leq I_2 \enspace .
\]
%
\noindent 
%
\begin{comment}
  \todo{check pre-order vs partial order} 
  % 
  Please note that~$\preceq$ is the Hoare order on power sets~\cite{AbramskyJung:1994aa}, which is a pre-order.
  % 
  For the anti-symmetry property, assume that~$t_1 \preceq t_2$ and~$t_2 \preceq t_1$: so for an~$I_{1} \in t_{1}$,
  there is $I_{2} \in t_{2}$ for which there is~$I_{3} \in t_{1}$ such that~$I_{1} \subseteq I_{2} \subseteq I_{3}$.
  % 
  But,~$t_{1}$ is maximal and, thus,~$I_{1} = I_{3} = I_{2}$. So,~$t_{1} = t_{2}$ and, thus,~$\preceq$ is a partial
  order.
\end{comment}

Similarly as for time intervals, the intuition for~$\preceq$ is that if a triple is true during the time points in all
the intervals in~$t_{2}$ and~$t_{1} \preceq t_{2}$, then, in particular, the triple is true at any time point in
intervals of~$t_{1}$. 
%
Essentially, if~$t_{1} \preceq t_{2}$ then a temporal triple~$\fuzzyg{\tau_{2}}{t_{2}}$ is true to a larger ``temporal
extent'' than the temporal triple~$\fuzzyg{\tau_{1}}{t_{1}}$. 

The partial order~$\preceq$ induces the following join ($\oplus$) operation on~$L$. Intuitively, if a triple is true
at~$t_{1}$ and also true at~$t_{2}$ then it will be true also for time points specified by~$t_{1} \oplus t_{2}$ (a kind
of union of time points).
%
As an example, if $\fuzzyg{\tau}{\{[2002,2005], [2008,2010]\}}$ and~$\fuzzyg{\tau}{\{[2004,2006], [2009,2012]\}}$ are
true then we expect that this is the same as saying that~$\fuzzyg{\tau}{\{[2002,2006], [2008,2012]\}}$ is true. 
%
The join operator will be defined in such a way that~$\{[2002,2005], [2008,2010]\} \oplus \{[2004,2006], [2009,2012]\} =
\{[2002,2006], [2008,2012]\}$.
%
Operationally, this means that~$t_{1} \oplus t_{2}$ will be obtained as follows: 
\begin{enumerate*}[label=(\roman*)]
\item take the union of the sets of intervals~$t = t_{1} \cup t_{2}$; and
\item join overlapping intervals in~$t$ until no more overlapping intervals can be obtained.
\end{enumerate*}
%
Formally,
%
\[
t_1 \oplus t_2 = \mathrm{infimum }\set{t \mid t \succeq t_{i}, i=1,2} \enspace .
\]
\nd It remains to define the~$\otimes$ over sets of intervals.
%
Intuitively, we would like to support inferences such that
%
from the triples~$\fuzzyg{\triple{a, \subclass, b}}{\{[2002,2005], [2008,2010]\}}$ and~$\fuzzyg{\triple{b, \subclass,
    c}}{\{[2004,2006], [2009,2012]\}}$, we can infer $\fuzzyg{\triple{a, \subclass, c}}{\{[2004,2005], [2009,2010]\}}$,
%
where~$\{[2002,2005], [2008,2010]\} \otimes \{[2004,2006]$, $[2009,2012]\} = \{[2004,2005], [2009,2010]\}$.
%
We get it by means of
% 
\[
t_1 \otimes t_2 = \mathrm{supremum }\set{t \mid t \preceq t_{i}, i=1,2 }  \enspace .
\]


\begin{example}[Temporal domain~$\otimes$]
  \label{exU1}
  %
  Using the following triples regarding another member of the Nightwish band
  %
  \[
  \begin{array}{l}
    \fuzzyg{\triple{\qname{}{NightwishMember}, \subclass, \qname{mo}{MusicArtist}}}{[1992,2012]}\\
    \fuzzyg{\triple{\qname{dbpedia}{Troy\mathunderscore{}Donockley}, \typeR, \qname{}{NightwishMember}}}{[1996, 1999]} \enspace ,
  \end{array} 
  \]
  % 
  we can infer that \vspace{-\abovedisplayskip}
  \[
    \fuzzyg{\triple{\qname{dbpedia}{Troy\mathunderscore{}Donockley}, \typeR, \qname{mo}{MusicArtist}}}{[1996, 1999]} \enspace ,
  \]
  where
  \(
  \{[1992,2012]\} \otimes \{[1996, 1999]\} = \{[1996, 1999]\}
  \).
  %
\end{example}




\subsubsection{The Fuzzy Domain}
To model fuzzy \ac{RDFS}~\cite{Straccia:2009aa} we may define the annotation domain as~$D_{[0,1]} = \tuple{[0,1], \max,
  \otimes, 0, 1}$ where~$\otimes$ is any continuous t-norm on~$[0,1]$.
%
\begin{example}[Fuzzy domain~$\otimes$]
  \label{exUSf}
  Adapting our running example to the fuzzy domain we can state the following:
  %
  Nightwish collaborators are partial members of the band (50\%), and since Troy is a Nightwish collaborator:
  %
  \[
  \begin{array}{l}
    \fuzzyg{\triple{\qname{}{NightwishCollaborator}, \subclass, \qname{}{NightwishMember}}}{0.5} \\
    \fuzzyg{\triple{\qname{dbpedia}{Troy\mathunderscore{}Donockley}, \typeR, \qname{}{NightwishCollaborator}}}{0.7}
  \end{array}
  \]
  \nd Then, \eg~under the product t-norm~$\otimes$, we can infer the following triple:
  \[
  \begin{array}{l}
    \fuzzyg{\triple{\qname{dbpedia}{Troy\mathunderscore{}Donockley}, \typeR, \qname{}{NightwishMember}}}{0.35}
  \end{array}
  \]
  %
\end{example}



\input{6-anql/provenance-domain-model}


\input{6-anql/deductive-system}



\input{6-anql/query-answering}




%%% Local Variables:
%%% mode: latex
%%% TeX-master: "../thesis"
%%% End:
