\chapter{A Secure RDF Data Integration Framework}
\label{cha:usecase}

In this chapter we go back to the \usecase presented in \cref{cha:introduction}, where we briefly mentioned the several
software applications that enterprises use to manage their business: interactions with clients in a Customer
Relationship Management (CRM) application, employee information in a Human Resources (HR) application, project
documentation and company policies in a Document Management System (DMS) and records of time spent working on projects
in a Timesheet System (TS).
%
Heterogeneity of the data formats from the different software applications is not the only problem. In fact, as much of
the information within the enterprise is highly sensitive, its integration could result in information leakage to
unauthorised individuals.

In this chapter we build on the languages presented in the previous chapters to automatically extract data and access
control information from the underlying databases and represent them as Annotated \ac{RDF} graphs, providing a holistic
view of data across the enterprise. 
%
This approach introduces a mechanism to enforce access control policies on the \ac{RDF} graph along with a flexible and
automatic way to represent and propagate the original access control policies.

In this chapter we define an annotation domain that models access control permissions as Annotated RDFS, specify the
high-level system architecture required to enforce access control by relying on SPARQL, and illustrate how domain
specific rules can be used to manage the access control annotations.
%
First we present some common access control related terms that we use in this chapter:
%
\begin{description}[noitemsep]
\item[Resources] denote the information to be protected;
  % 
\item[Users] represent individuals requesting access to resources;
  % 
\item[Groups] are collections of users with common features (\eg~contributors, supervisors, and management);
  % 
\item[Roles] are commonly used to assign access rights to a set of individuals and groups, for example by department
  (\eg~human resources, sales and marketing) or task (\eg~insurance claim processing, reporting and invoicing).
  % 
\end{description}

\section{The Access Control Annotation Domain}
\label{sec:access-contr-doma}

In this section we formalise our access control annotation domain, following the definitions presented in
\cref{sec:rdfs-annot-doma}.  We start by defining the entities and annotation values and then present the~$\otimes$
and~$\oplus$ domain operations. Finally, we briefly describe the implementation of the presented annotation domain.


\subsection{Entities and Annotations}
\label{sec:modell-access-contr}

\input{7-usecase/domain-modelling-negation}


\subsection{Domain Implementation}
\label{sec:impl-annot-doma}

\input{7-usecase/domain-implementation-negation}



\input{7-usecase/framework-evaluation}


\section{Related Work}
\label{sec:related-work}

\begin{comment}
  The use of partially ordered sets for access control has previously been considered in the domain of information flow
  in computer systems, mostly in response to the introduction of multi-user systems~\cite{Denning:1976aa}.  Although
  focusing on the question of information passing between arbitrary objects of a computer system (e.g.\ programs,
  processes or files), each process also includes a \emph{clearance class} that limits the security classes the process
  can read and write.
\end{comment}


The topic of access control has been long studied in relational databases and the approach of enforcing the access
policies by query rewriting was also considered for the Quel query language by~\citet{StonebrakerWong:1974aa}.  However,
the presented system does not rely on annotating the relational data but rather access control is specified using
constraints over the user credentials, which are then included in the rewritten query.
%
An overview of common issues, existing models and languages for access control is provided
by~\citet{VimercatiSamaratiJajodia:2005aa}.  


For the Semantic Web, well known policy languages such as KAoS~\cite{Bradshaw1997}, Rei~\cite{Kagal2003} and
PROTUNE~\cite{Bonatti2009}.  Although such languages enable policy specification using \ac{RDF} and \ac{OWL}, in their
current form they do not support reasoning based on \ac{RDF} data.  These policy languages are complimentary to our work
as they can be mapped to our annotations using rules.


\citet{DietzoldAuer2006aa} describe the requirements an \ac{RDF} store needs from a Semantic Wiki perspective.  Apart
from the necessary requirements on efficiency and scalability, the authors refer the need for access control on a triple
level and the need to integrate the structure of the organisation in the access control methods.  The described system
relies on a query engine (SPARQL is mentioned but no details are given) and a rule processor to decide the access
control enforcement at query time.
%
The system we propose in this chapter caters for both of these requirements and also integrates the access control into
the annotation query language.

\citet{HollenbachPresbreyBerners-Lee:2005aa} present the possibility of maintaing metadata on the \ac{RDF} data to
enforce access control and discuss, as possible extensions of their model, some of the work presented here, \eg~using
rules for specifying access control.  Providing access control on a resource level is also left as an open question, one
we are tackling by the specification of rules.  The extension of SPARQL is not considered.


Similar access control annotations are considered attached to axioms in an ontology
by~\citet{KnechtelStuckenschmidt:2010aa} and~\citet{BaaderKnechtelPenaloza:2009aa} and are used to allow access to
subsets of the ontology to specific users and also apply such annotations to the problem of determining the minimal set
of axioms that are necessary to support a certain conclusion.  Although the setting is different to the one presented in
this chapter, some of the algorithms for efficient annotation calculation may be ported to our modelling.


Some work on extending query languages was presented by~\citet{AbelCoiHenze2007aa}, however this work pre-dates the
SPARQL query language.  In a similar fashion to the work proposed in this chapter, their policy enforcement is also done
by a query rewriting step, however their query rewriting does not consists of including the user credentials but rather
replicating the access policies within the query.  
%
They also consider access control policies that both grant and restrict access to data.

                 
\section{Conclusion}

In this chapter we proposed an Access Control model, which can be used to protect \ac{RDF} data and demonstrate how a
combination of Annotated RDFS and SPARQL can be used to control access to integrated enterprise data.
%
This model is based on the Annotated RDFS framework presented in \cref{cha:anql} and attaches the access control
information on a triple basis, \ie~each \ac{RDF} triple can contain different annotation values.  
%
This solution provides a flexible representation method for the access control annotations, based on propositional
formul\ae{} to define which entities have access to the triple.
%
However, when considering large number of triples, challenges arise with respect to optimal access control policy
administration.  To tackle this issue we propose permission management through specifying domain-specific inference
rules for the annotation domain.
%
We also suggest a possible implementation structure for a framework to enforce the access control based on rewriting a
SPARQL query into an AnQL query.





%%% Local Variables:
%%% mode: latex
%%% mode: flyspell
%%% mode: reftex
%%% TeX-master: "../thesis"
%%% End:
