\chapter{Query Languages}
\label{cha:query-languages}

Query languages allow users to select and transform data from large sources.  The ability to select only relevant data
is an essential feature to minimise serialisation and communication overheads, especially when we consider the
transmission of data over the Web.
%

Due to the specific characteristics of each data model, query languages are usually tailored to work with a single data
model.
%
For the data models presented in the previous section, the respective query languages are \acs{SQL}, XQuery and SPARQL,
for which we will give an overview next.
%
We also present the closely related \acs{XSLT} transformation language for \ac{XML} data.
%

In \cref{fig:data-models-overview} we presented a high-level overview of the available languages for each data model.
Data and schema manipulation languages are widely available for relational data, for \ac{XML} an update language has
been recently standardised~\cite{RobieChamberlinDyck:2011aa} while for \ac{RDF} data this feature is only included in
the upcoming version of SPARQL~1.1~\cite{GearonPassantPolleres:2012aa}.

In the following sections we start by presenting a short overview of the possible forms of querying relational
databases, including the \ac{SQL} query language, before turning to the different query languages for \ac{XML} in
\cref{sec:prelim-xquery}.  In this section we again present a short overview of the XPath and \ac{XSLT} languages and
then focus in more detail on the XQuery language, which will be the basis for the XSPARQL language in
\cref{cha:xsparql}.
%
Finally, \cref{sec:sparql-preliminaries} provides a detailed description of the SPARQL query language for
\ac{RDF} (which we will extend in \cref{cha:anql}).


\input{3-query-languages/1-sql}

\input{3-query-languages/2-querying-xml}

\input{3-query-languages/3-sparql }



\section{Conclusion}
\label{sec:conclusion-query-languages}

This chapter introduced the \ac{SQL}, XQuery, and SPARQL query languages that allow to access data in the formats
presented in \cref{cha:data-models}.
%
Each query language focuses on a specific data model, namely \ac{SQL} for relational data, XQuery for \ac{XML} data, and
SPARQL for \ac{RDF} data.  For \ac{XML}, we briefly presented the \ac{XPath} and \ac{XSLT} languages, which are closely
related to XQuery.

We briefly introduced the syntax and semantics of each language, with special focus on XQuery and SPARQL, which will be
used in the next chapters to define the novel transformation language, called XSPARQL, and the extension of SPARQL
towards querying meta-information.
%
XSPARQL integrates the \ac{SQL}, XQuery, and SPARQL query languages presented in this chapter, thus allowing to combine
data from the different data models.


%%% Local Variables:
%%% mode: latex
%%% mode: flyspell
%%% mode: reftex
%%% TeX-master: "../thesis"
%%% End:
