\chapter{XSPARQL Evaluation and Optimisations}
\label{cha:optimisations}

\def\plotsDir{5-optimisations/benchmarks/results/2012.08.13-10.06/plots/}


This chapter describes our prototype implementation of the XSPARQL language presented in the previous chapter.
%
We then describe a benchmark suite that will be used to evaluate our XSPARQL implementation.  This benchmark is based on
a widely used \ac{XML} benchmark suite (XMark), and extends it to cater for \ac{XML} and \ac{RDF} data. 
%
The experimental evaluation will show that, in our current implementation, nested queries with an inner \SparqlForClause
present the highest overhead when compared to their XQuery counterpart.

To tackle this issue, \cref{sec:optimisation} details different possible approaches for evaluating nested queries
in our prototype and compares these approaches regarding their evaluation times.  
%
In \cref{sec:related-work-optimisations} we present an overview of work related to the optimisations presented in
this chapter.


\input{5-optimisations/4-xsparql-implementation}

\input{5-optimisations/xmarkRDF-description}

\input{5-optimisations/6-xsparql-optimisations}


\section{Related Work}
\label{sec:related-work-optimisations}


In this related work section we present a comparison of different approaches for nesting in SPARQL, including the
proposal from the new version of SPARQL, SPARQL~1.1.

The new version of SPARQL was presented in \cref{sec:sparql-11} and introduces many new features that we already
support in XSPARQL.
%
Most notably:
%
\begin{enumerate*}[label=(\roman*), noitemsep]
\item construction of values in \SELECT expressions;
\item variable assignment;
\item remote endpoint querying; and
\item subqueries (or query nesting).
\end{enumerate*}
%
The value construction and variable assignment features behave similar to the features in XSPARQL. 
%
However, the XSPARQL language provides a convenient syntax for mapping between different \ac{RDF} vocabularies, \eg~by
generating blank node labels, a task that can otherwise be cumbersome even in SPARQL~1.1.
% 
\begin{query}
  \lstinputlisting[language=XSPARQL]{0-data+queries/rdf2rdf-bands.xsparql}%
  \caption{Transformation between RDF representations in XSPARQL}%
  \label{qr:rdf2rdf-xsparql}%
\end{query}
%
\begin{query}
  \lstinputlisting[language=XSPARQL]{0-data+queries/bnode.sparql}%
  \caption{Transformation between RDF representations in SPARQL~1.1}%
  \label{qr:rdf2rdf-sparql}%
\end{query}
%
\begin{example}[Translation between RDF vocabularies]
  \Cref{qr:rdf2rdf-xsparql} presents a simple transformation between different \ac{RDF} vocabularies: from using
  \acp{URI} to identify bands to blank nodes, where we create a blank node that identifies each band and replicate its
  members.
  % 
  A more involved but equivalent query written in SPARQL~1.1 is presented in~\cref{qr:rdf2rdf-sparql}.
  % 
\end{example}
%
Subqueries (along with the special form of subquery that is the remote endpoint query) still presents noteworthy
differences between our approach in XSPARQL and the approach proposed in SPARQL~1.1.
%
These differences, also briefly highlighted in \cref{sec:external-endpoint}, are mostly related to the evaluation
method of the different languages, while SPARQL follows a bottom up approach, XQuery and thus XSPARQL follow a top-down
approach.
%
In SPARQL, the subqueries are evaluated and the produced bindings are then merged with the bindings from the outer query and
such subqueries must be executed over the same dataset as the outer query, \ie~\FROM and \FROMNAMED clauses are not
allowed in subqueries.
%
However, this evaluation method prevents the reuse of variables declared in the inner query.
%
The different evaluation models cater for different and complementary \usecases.  While the method followed by \ac{SQL}
and SPARQL is suitable for parallel and distributed computing, the model followed by XQuery (and thus XSPARQL), allowing
to inject values into the inner expressions, can be a necessary feature.

Also related to our nested queries optimisation,~\citet{AnglesGutierrez:2010aa} presented initial work on an extension
of SPARQL that caters for nested queries and presented preliminary equivalences between types of nested queries with the
aim of determining if query unnesting can be successfully applicable.
%
The same authors then extended this work in~\citet{AnglesGutierrez:2011aa}, where they consider different forms of
nesting, namely nesting in \FROM clauses, nesting in graph patterns, and nesting in filter expressions.
%
In XSPARQL, we easily support the nesting in \FROM clauses by assigning the result of a \CONSTRUCT query to a variable
and reusing this variable in a \FROM clause.  Nesting in graph patterns can also be simulated in XSPARQL by using the
standard XQuery nesting of expressions, in this case nesting \SparqlForClause{s}.  
%
Regarding nesting in \FILTER expressions, although possible to implement in XSPARQL, this would require processing of
the results from the SPARQL query in XSPARQL.




The presented approaches for query rewriting applied to XSPARQL nested queries is similar to already known optimisations
from the relational databases realm and also presented for XQuery queries by~\citet{MayHelmerMoerkotte:2003aa}.  
%
This work proposes unnesting equivalences that, while maintaining the element order, provide significant performance
gains.  For the original XMark nested queries~$q_{8}$ and~$q_{11}$, we can consider using the equivalences described for
the ``Grouping and Aggregation'' unnesting equivalences, whereas~$q_{9}$ and~$q_{10}$ rely on the ``Grouping''
equivalences.  Hence, the nested-loop rewriting of the queries we present replicates the unnested query plans for these
optimisations.




\section{Conclusion}
\label{sec:conclusions-optimisations}

In this chapter we presented our implementation of the XSPARQL language presented in \cref{cha:xsparql}.  Our
implementation attempts to reuse \emph{off-the-shelf} components, where we translate each XSPARQL query into an XQuery
containing interleaved calls to a \ac{SQL} and/or SPARQL engine.
%
The benchmark evaluation of our implementation has shown that nested queries incur the highest evaluation overhead and
thus we presented different rewritings that aim at reducing this overhead.

%
The presented optimisations are based on reordering the expressions in the XQuery rewriting to minimise the number of
calls to the SPARQL endpoint or based on performing a more complex SPARQL query that takes care of joining the
variables.
%
The benchmarks that were carried out to determine the impact of our optimisations have shown encouraging results for
nested expressions whose inner expressions access \ac{RDF} data, hinting at a large potential for optimisations in
XSPARQL.
%
However, similar rewritings do not produce the excepted improvements for nested expressions that access relational
data.   This indicates that different optimisations need to be considered for accessing relational data.

Among the rewriting strategies presented in this chapter and on our test data, pushing joins into a SPARQL engine appeared
the most promising strategy.
%
Our benchmark results showed that our optimisations are not only specific to XSPARQL having also improved the response
times of the SPARQL2XQuery system to which we compared XSPARQL.

%
Also according to our benchmarks, encoding SPARQL in XQuery seems a viable option -- assuming that we have access to the
\ac{RDF} dataset beforehand -- that would allow to compile XSPARQL to pure XQuery without the use of a separate SPARQL
engine.




%%% Local Variables:
%%% mode: latex
%%% mode: flyspell
%%% mode: reftex
%%% TeX-master: "../thesis"
%%% End:
