
\chapter*{Acknowledgements\markboth{Acknowledgements}{Acknowledgements}}

First of all I would like to thank Axel and Stefan for all their help, support, and directions provided during my Ph.D.
%
Thanks to my examiners Claudio Guti{\'e}rrez and Siegfried Handschuh for their time.

Also a big thanks to all of my co-authors and co-workers for countless hours spent tackling interesting problems.
%
Notably the persons I was closely working with: Stefan Bischof, Gergely Luk{\'a}csy, Antoine Zimmermann, Sabrina
Kirrane, Umberto Straccia, Thomas Krennwallner, and other URQ members.

I would like to thank all the people that helped in reviewing and proofreading this thesis: Sabrina Kirrane, Deirdre
Lee, Antoine Zimmermann, Alexandre Passant, Stefan Bischof, and Aidan Hogan.  Especially thanks to Aidan and Sabrina for
all their invaluable comments and their patience for my incessant nagging.
%
Thanks also to J{\"u}rgen: it was great to have someone close to share the agony of thesis writing.

For the times when we needed some distraction from hard work the Foosball (thanks to the UDI2 members that supplied the
table) or Magic The Gathering crowd was always close by!


Above all, I would like to thank my family and give credit to my partner Ana, for all the support and always
understanding the (many) times when I had to prioritise work.


\quotationpage{%
  To truly know something, you must become it. The trick is to not lose yourself in the process.%
}{%Flavour text,%
  ``Thirst for Knowledge'', Magic the Gathering card% \# 68, Expansion Duel Decks: Elspeth vs. Tezzeret
}




%%% Local Variables:
%%% mode: latex
%%% mode: flyspell
%%% mode: reftex
%%% TeX-master: "thesis"
%%% End:
